---
title: "A Modern Semi-Structural Monetary Policy Model for Brazil"
subtitle: "A Monthly MPP Adaptation Inspired by Bogdanski, Tombini & Werlang (2000)"
author: "Inspired by the BCB Small-Scale Model (MPP) — Bogdanski, Tombini & Werlang (2000)"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
    code_folding: show
    css: |
      body { font-family: 'Georgia', serif; font-size: 15px; line-height: 1.7; color: #1a1a2e; }
      h1, h2, h3 { font-family: 'Trebuchet MS', sans-serif; color: #16213e; }
      h1.title { font-size: 2.2em; color: #0f3460; border-bottom: 3px solid #e94560; padding-bottom: 12px; }
      .subtitle { color: #457b9d; font-style: italic; font-size: 1.1em; }
      .highlight-box {
        background: #eaf4fb;
        border-left: 5px solid #0f3460;
        padding: 14px 18px;
        margin: 18px 0;
        border-radius: 4px;
      }
      .warning-box {
        background: #fff8e1;
        border-left: 5px solid #f59e0b;
        padding: 14px 18px;
        margin: 18px 0;
        border-radius: 4px;
      }
      blockquote {
        border-left: 4px solid #e94560;
        color: #555;
        font-style: italic;
        padding-left: 14px;
      }
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  message    = FALSE,
  warning    = FALSE,
  fig.align  = "center",
  dpi        = 150
)

pkgs <- c("tidyverse", "mFilter", "dynlm", "sandwich", "lmtest",
          "ggplot2", "plotly", "knitr", "kableExtra", "scales")

invisible(lapply(pkgs, function(p) {
  if (!requireNamespace(p, quietly = TRUE)) install.packages(p, quiet = TRUE)
  library(p, character.only = TRUE)
}))

# ── Plotly layout defaults ─────────────────────────────────────────────────────
plt_layout <- list(
  plot_bgcolor  = "white",
  paper_bgcolor = "white",
  font          = list(family = "Georgia, serif", size = 12),
  margin        = list(l = 60, r = 30, t = 70, b = 50),
  legend        = list(orientation = "h", y = -0.18, font = list(size = 11)),
  hoverlabel    = list(bgcolor = "white", font = list(size = 12))
)

plt_cfg <- list(displayModeBar = TRUE, displaylogo = FALSE,
                modeBarButtonsToRemove = list("lasso2d", "select2d"))

clr_main  <- "#0f3460"
clr_red   <- "#e94560"
clr_blue  <- "#1a7abf"
clr_teal  <- "#2ab7ca"
clr_grey  <- "#888888"

# plotly horizontal line helper (add_hline is a ggplot2 function, not plotly)
hline <- function(y, color = "#888888", dash = "dot", width = 1) {
  list(type = "line",
       x0 = 0, x1 = 1, xref = "paper",
       y0 = y, y1 = y,
       line = list(color = color, dash = dash, width = width))
}
```

---

::::: {.warning-box}
**Methodological positioning — read before proceeding.**

This document does **not** replicate the operational model used by the Central Bank of Brazil in 1999. It develops a **modern, monthly, semi-structural adaptation** inspired by the four-block architecture of Bogdanski, Tombini & Werlang (2000): aggregate demand (IS), inflation dynamics (Phillips), exchange-rate transmission (UIP), and policy rules (Taylor).

Key differences from the original MPP:

- **Frequency**: the original model operated quarterly; this adaptation runs on **monthly data** (IBC-Br as activity proxy, Focus survey expectations, EMBI+ country risk).
- **Estimation**: the original combined calibration and judgmental adjustment; this note uses **OLS with HAC standard errors** without formally imposing all structural restrictions.
- **Neutral rate**: the original imposed $r^* = g^*$ as a hard constraint; this adaptation estimates a **statistical trend of the ex-ante real rate** as a proxy — it should not be interpreted as a structural neutral rate.
- **Scope**: the original focused on conditional inflation forecasts under alternative SELIC paths; this adaptation also benchmarks the structural model against machine learning methods.

Use this note as a **pedagogical reference** for the monetary policy transmission mechanism in Brazil, not as a replication of the BCB's internal toolkit.
:::::

---

# Motivation: Why Small-Scale Models?

The inflation-targeting regime requires the central bank to adopt a **forward-looking** stance: monetary policy decisions affect inflation with considerable lags. As Greenspan put it:

> *"Implicit in any monetary policy action or inaction is an expectation of how the future will unfold, that is, a **forecast**."*

The MPP occupies a strategic position in the model hierarchy: small enough to explain to a policy committee, structured enough to simulate alternative monetary policy scenarios.

```{r model-hierarchy, echo=FALSE}
df_hier <- data.frame(
  model    = c("Leading\nIndicators", "VAR / ARMA", "MPP\n(Semi-struct.)", "Full DSGE"),
  x        = 1:4,
  transpar = c(90, 70, 55, 20),
  forecast = c(30, 60, 70, 85),
  tooltip  = c(
    "Leading Indicators\nTransparency: High\nSimulation capacity: Low",
    "VAR / ARMA\nTransparency: Medium\nSimulation capacity: Medium",
    "MPP (Semi-structural)\nTransparency: Medium\nSimulation capacity: High",
    "Full DSGE\nTransparency: Low\nSimulation capacity: Very High"
  )
)

plot_ly(df_hier,
        x    = ~transpar, y = ~forecast,
        type = "scatter", mode = "markers+text",
        marker = list(size = 22, opacity = 0.85,
                      color = c(clr_red, clr_main, clr_blue, clr_teal)),
        text      = ~model,
        textposition = "top center",
        hovertext = ~tooltip,
        hoverinfo = "text",
        showlegend = FALSE) %>%
  layout(
    title = list(
      text = "<b>Macroeconomic Model Hierarchy</b><br><sub>Transparency vs. richness of mechanisms trade-off</sub>",
      font = list(size = 14, color = clr_main)
    ),
    xaxis = list(title = "← Complexity / Micro-foundations",
                 autorange = "reversed",
                 showticklabels = FALSE, zeroline = FALSE),
    yaxis = list(title = "Policy simulation capacity →",
                 showticklabels = FALSE, zeroline = FALSE),
    plot_bgcolor  = "white",
    paper_bgcolor = "white",
    font = plt_layout$font,
    margin = plt_layout$margin,
    hoverlabel = plt_layout$hoverlabel
  ) %>%
  config(displayModeBar = FALSE)
```

---

# Model Structure

The MPP consists of **four main blocks**. The diagram below shows their articulation (static — hover on subsequent charts for detail):

```{r block-diagram, echo=FALSE, fig.height=3.8, fig.width=8}
library(ggplot2)   # static diagram — plotly adds no value here

nodes <- data.frame(
  label = c("IS Curve\n(Aggregate Demand)",
            "Phillips Curve\n(Aggregate Supply)",
            "Uncovered\nInterest Parity",
            "Monetary\nPolicy Rule",
            "INFLATION"),
  x = c(1, 3, 3, 1, 2),
  y = c(2, 2, 0, 0, 1),
  fill = c(clr_main, clr_main, clr_main, clr_red, clr_teal)
)
edges <- data.frame(
  x0 = c(1,1,3,3,1), y0 = c(2,0,0,2,2),
  x1 = c(2,2,2,2,3), y1 = c(1,1,1,1,2)
)
ggplot() +
  geom_segment(data = edges, aes(x=x0,y=y0,xend=x1,yend=y1),
               arrow = arrow(length=unit(0.2,"cm"), type="closed"),
               color = "#457b9d", linewidth = 0.8) +
  geom_label(data = nodes, aes(x=x, y=y, label=label),
             fill = nodes$fill, color = "white", size = 3.2, fontface = "bold",
             label.padding = unit(0.4,"lines"), label.r = unit(0.15,"lines")) +
  xlim(0.3,3.7) + ylim(-0.4,2.6) +
  theme_void() +
  labs(title = "MPP Block Diagram",
       caption = "Source: Bogdanski, Tombini & Werlang (2000), Figure 1") +
  theme(plot.title   = element_text(face="bold", color=clr_main, size=13),
        plot.caption = element_text(color="#888", size=9))
```

---

# Block I — IS Curve (Aggregate Demand)

## Theoretical Specification

$$\boxed{\tilde{y}_t = \beta_0 + \beta_1 \tilde{y}_{t-1} + \beta_2 \tilde{y}_{t-2} + \beta_3 r_{t-1} + \varepsilon^d_t}$$

where $\tilde{y}_t = \ln(Y_t / Y^*_t)$ is the output gap and $r_t$ is the ex-ante real interest rate.

::::: {.warning-box}
**Neutral rate note.** The original MPP imposed $r^* = g^*$ as a hard structural restriction. This adaptation estimates a *statistical trend of the ex-ante real rate* as a proxy. That average reflects BCB reactions to shocks and crises — it is not automatically the equilibrium rate consistent with potential growth.
:::::

## Output Gap and HP Filter

```{r hp-gap}
set.seed(42)
n     <- 120
trend <- cumsum(c(0.005, rep(0.006, n-1) + rnorm(n-1, 0, 0.001)))
cycle_v <- numeric(n)
for(t in 2:n) cycle_v[t] <- 0.65 * cycle_v[t-1] + rnorm(1, 0, 0.01)

idx_gdp <- 100 + trend + cycle_v
dates   <- seq(as.Date("1995-01-01"), by = "quarter", length.out = n)

df_gdp <- data.frame(date = dates, idx_gdp = idx_gdp)
hp_result        <- mFilter::hpfilter(df_gdp$idx_gdp, freq = 1600)
df_gdp$potential <- hp_result$trend
df_gdp$gap       <- hp_result$cycle

# ── Chart 1: observed vs potential ─────────────────────────────────────────────
p_gdp <- plot_ly(df_gdp, x = ~date) %>%
  add_lines(y = ~idx_gdp,
            name = "Observed GDP",
            line = list(color = clr_main, width = 2),
            hovertemplate = "%{x|%Y-%m}<br>Observed: %{y:.2f}<extra></extra>") %>%
  add_lines(y = ~potential,
            name = "Potential GDP (HP)",
            line = list(color = clr_red, width = 2, dash = "dash"),
            hovertemplate = "%{x|%Y-%m}<br>Potential: %{y:.2f}<extra></extra>") %>%
  layout(
    title  = list(text = "<b>Observed GDP vs. Potential Output</b><br><sub>Hodrick-Prescott filter (λ = 1,600) — simulated data</sub>",
                  font = list(size = 13, color = clr_main)),
    xaxis  = list(title = ""),
    yaxis  = list(title = "Index (simulated)"),
    hovermode    = "x unified",
    plot_bgcolor = "white", paper_bgcolor = "white",
    legend = plt_layout$legend, font = plt_layout$font,
    margin = plt_layout$margin, hoverlabel = plt_layout$hoverlabel
  )

# ── Chart 2: output gap ─────────────────────────────────────────────────────────
p_gap <- plot_ly(df_gdp, x = ~date) %>%
  add_ribbons(ymin = ~pmin(gap, 0), ymax = ~pmax(gap, 0),
              fillcolor = paste0(clr_main, "30"),
              line = list(color = "transparent"),
              name = "Gap area", showlegend = FALSE,
              hoverinfo = "skip") %>%
  add_lines(y = ~gap,
            name = "Output gap",
            line = list(color = clr_main, width = 2),
            hovertemplate = "%{x|%Y-%m}<br>Gap: %{y:.4f}<extra></extra>") %>%
  layout(shapes = list(hline(0, clr_grey, "dot"))) %>%
  layout(
    title  = list(text = "<b>Output Gap</b><br><sub>Deviation from HP trend (index units)</sub>",
                  font = list(size = 13, color = clr_main)),
    xaxis  = list(title = ""),
    yaxis  = list(title = "Gap (index units)"),
    hovermode    = "x unified",
    plot_bgcolor = "white", paper_bgcolor = "white",
    legend = plt_layout$legend, font = plt_layout$font,
    margin = plt_layout$margin, hoverlabel = plt_layout$hoverlabel
  )

subplot(p_gdp, p_gap, nrows = 2, shareX = TRUE, titleY = TRUE) %>%
  layout(hovermode = "x unified") %>%
  config(plt_cfg)
```

## IS Estimation

```{r is-estimation}
set.seed(7)
nominal_rate  <- 0.18 + 0.3 * (-df_gdp$gap) + rnorm(n, 0, 0.03)
inflation_sim <- 0.06 + 0.5 * df_gdp$gap    + rnorm(n, 0, 0.015)
real_rate     <- nominal_rate - inflation_sim

df_is <- df_gdp %>%
  mutate(y = gap, y_l1 = lag(gap,1), y_l2 = lag(gap,2), r_l1 = lag(real_rate,1)) %>%
  drop_na()

model_is    <- lm(y ~ y_l1 + y_l2 + r_l1, data = df_is)
errors_hac  <- sandwich::NeweyWest(model_is, lag = 4, prewhite = FALSE)
coeftest_is <- lmtest::coeftest(model_is, vcov = errors_hac)

knitr::kable(
  broom::tidy(coeftest_is) %>%
    mutate(term = c("β₀ (intercept)","β₁ (ỹ_{t-1})","β₂ (ỹ_{t-2})","β₃ (r_{t-1})"),
           across(where(is.numeric), ~round(.x,4))) %>%
    rename(Parameter=term, Estimate=estimate,
           `Std. Error`=std.error, `t-stat`=statistic, `p-value`=p.value),
  caption = "**IS Curve — OLS with HAC errors (Newey-West, 4 lags)**"
) %>%
  kableExtra::kable_styling(bootstrap_options=c("striped","hover","condensed"), full_width=FALSE)
```

```{r is-diagnostics}
df_is$fitted <- fitted(model_is)
df_is$resid  <- residuals(model_is)

p_fit <- plot_ly(df_is, x = ~date) %>%
  add_lines(y = ~y*100, name = "Observed",
            line = list(color = clr_main, width = 1.8),
            hovertemplate = "%{x|%Y-%m}<br>Obs: %{y:.3f}<extra></extra>") %>%
  add_lines(y = ~fitted*100, name = "Fitted",
            line = list(color = clr_red, width = 1.8, dash = "dash"),
            hovertemplate = "%{x|%Y-%m}<br>Fitted: %{y:.3f}<extra></extra>") %>%
  layout(title = list(text = "<b>Observed vs. fitted output gap</b>",
                      font = list(size = 12, color = clr_main)),
         xaxis = list(title = ""), yaxis = list(title = "Gap × 100"),
         hovermode = "x unified", plot_bgcolor="white", paper_bgcolor="white",
         legend = plt_layout$legend, font = plt_layout$font,
         margin = plt_layout$margin, hoverlabel = plt_layout$hoverlabel)

p_res <- plot_ly(df_is, x = ~date) %>%
  add_bars(y = ~resid*100, name = "Residual",
           marker = list(color = paste0(clr_blue, "99")),
           hovertemplate = "%{x|%Y-%m}<br>Residual: %{y:.4f}<extra></extra>") %>%
  layout(shapes = list(hline(0, clr_grey, "dot"))) %>%
  layout(title = list(text = "<b>IS Curve residuals</b>",
                      font = list(size = 12, color = clr_main)),
         xaxis = list(title = ""), yaxis = list(title = "Residual"),
         hovermode = "x unified", plot_bgcolor="white", paper_bgcolor="white",
         legend = plt_layout$legend, font = plt_layout$font,
         margin = plt_layout$margin, hoverlabel = plt_layout$hoverlabel)

subplot(p_fit, p_res, nrows = 2, shareX = TRUE, titleY = TRUE) %>%
  layout(hovermode = "x unified") %>%
  config(plt_cfg)
```

---

# Block II — Phillips Curve (Aggregate Supply)

## Hybrid Specification

$$\boxed{\pi_t = \underbrace{\alpha_1 \pi_{t-1} + \alpha_2 \pi_{t-2}}_{\text{inertia}} + \underbrace{\alpha_{fwd}\,\mathbb{E}_t[\pi_{t+1}]}_{\text{expectations}} + \alpha_3 \tilde{y}_{t-1} + \alpha_4 \Delta \ln E_t + \varepsilon_t}$$

Long-run restriction: $\alpha_1 + \alpha_2 + \alpha_{fwd} = 1$.

::::: {.warning-box}
**OLS caveat.** This adaptation does not formally impose the long-run restriction. Additionally, 12-month accumulated CPI creates mechanical autocorrelation across observations; and Focus expectations should be verified as genuinely ex-ante to avoid look-ahead bias.
:::::

## Simulation and Pass-Through

```{r phillips-simulation}
set.seed(21)
# α₁ + α₂ + α_fwd = 0.45 + 0.20 + 0.35 = 1.00 ✓ (long-run neutrality)
# α₃ (output gap) and α₄ (FX) are separate — not part of the inflation restriction
alpha1 <- 0.45; alpha2 <- 0.20; alpha_fwd <- 0.35
alpha3 <- 0.25; alpha4 <- 0.12
pi_star <- 0.045

delta_e <- rnorm(n, 0, 0.04)
eps_pc  <- rnorm(n, 0, 0.008)
pi <- numeric(n); pi[1:3] <- 0.06
for (t in 4:n) {
  pi_fwd <- pi[t-1] + rnorm(1, 0, 0.005)
  pi[t]  <- alpha1*pi[t-1] + alpha2*pi[t-2] + alpha_fwd*pi_fwd +
            alpha3*df_gdp$gap[t] + alpha4*delta_e[t] + eps_pc[t]
  pi[t]  <- max(pi[t], -0.02)
}

df_pc <- data.frame(date = dates, inflation = pi, gap = df_gdp$gap, delta_e = delta_e)

# ── Panel 1: simulated inflation ────────────────────────────────────────────────
p_pi <- plot_ly(df_pc, x = ~date) %>%
  add_lines(y = ~inflation*100, name = "Inflation",
            line = list(color = clr_red, width = 2),
            hovertemplate = "%{x|%Y-%m}<br>Inflation: %{y:.2f}%<extra></extra>") %>%
  layout(shapes = list(hline(pi_star*100, clr_main, "dash", 1.5))) %>%
  layout(
    title = list(text = "<b>Simulated inflation — Hybrid Phillips Curve</b><br><sub>α₁ + α₂ + α_fwd = 1.00 (long-run neutrality)</sub>",
                 font = list(size = 12, color = clr_main)),
    xaxis = list(title = ""), yaxis = list(title = "CPI (% y/y)", ticksuffix = "%"),
    hovermode = "x unified", plot_bgcolor="white", paper_bgcolor="white",
    legend = plt_layout$legend, font = plt_layout$font,
    margin = plt_layout$margin, hoverlabel = plt_layout$hoverlabel
  )

# ── Panel 2: pass-through specifications ───────────────────────────────────────
E_seq  <- seq(1.5, 6.5, length.out = 100)
df_pt  <- data.frame(
  E         = E_seq,
  Constant  = 0.12,
  Quadratic = 0.08 + 0.04 * (log(E_seq))^2,
  Level     = pmax(0.05, 0.22 - 0.03 * log(E_seq))
)

p_pt <- plot_ly(df_pt, x = ~E) %>%
  add_lines(y = ~Constant,  name = "Constant",
            line = list(color = clr_main, width = 2),
            hovertemplate = "BRL/USD: %{x:.2f}<br>α₄ (const): %{y:.3f}<extra></extra>") %>%
  add_lines(y = ~Quadratic, name = "Quadratic in Δe",
            line = list(color = clr_red, width = 2),
            hovertemplate = "BRL/USD: %{x:.2f}<br>α₄ (quad): %{y:.3f}<extra></extra>") %>%
  add_lines(y = ~Level,     name = "Level-dependent",
            line = list(color = clr_teal, width = 2),
            hovertemplate = "BRL/USD: %{x:.2f}<br>α₄ (level): %{y:.3f}<extra></extra>") %>%
  layout(
    title = list(text = "<b>Exchange rate pass-through: alternative specifications</b><br><sub>Goldfajn & Werlang (1999)</sub>",
                 font = list(size = 12, color = clr_main)),
    xaxis = list(title = "Exchange Rate (BRL/USD)"),
    yaxis = list(title = "Coefficient α₄"),
    hovermode = "x", plot_bgcolor="white", paper_bgcolor="white",
    legend = plt_layout$legend, font = plt_layout$font,
    margin = plt_layout$margin, hoverlabel = plt_layout$hoverlabel
  )

subplot(p_pi, p_pt, nrows = 2, shareX = FALSE, titleY = TRUE) %>%
  config(plt_cfg)
```

---

# Block III — Uncovered Interest Parity (UIP)

## Corrected Sign Convention

With $e_t = \ln(\text{BRL/USD})$, the UIP in first differences is:

$$\boxed{\Delta e_t = -\alpha_i \Delta i_t + \alpha_f \Delta i^*_t + \alpha_x \Delta x_t + \eta_t}$$

with $\alpha_i > 0,\ \alpha_f > 0,\ \alpha_x > 0$.

::::: {.warning-box}
**Sign error to avoid.** The domestic SELIC rate must enter with a **negative** coefficient. Under $e_t = \ln(\text{BRL/USD})$, a COPOM hike appreciates the BRL (reduces $e_t$). Writing it with a positive sign inverts the entire exchange-rate transmission channel.
:::::

```{r uip-simulation}
set.seed(99)
i_ext <- 0.055
x_risk <- numeric(n); e_log <- numeric(n)
x_risk[1] <- 300; e_log[1] <- log(3.5)

for(t in 2:n) {
  x_risk[t] <- max(0.7*x_risk[t-1] + rnorm(1,0,25) - 15*(t>60), 50)
  selic_t   <- 0.12 - 0.02*(t/n) + rnorm(1,0,0.015)
  # UIP: higher domestic rate → BRL appreciation → negative sign
  e_log[t]  <- e_log[t-1] - (selic_t - i_ext - x_risk[t]/10000)/4 + rnorm(1,0,0.025)
}

df_uip <- data.frame(date = dates, fx = exp(e_log), risk = x_risk)

p_e <- plot_ly(df_uip, x = ~date) %>%
  add_lines(y = ~fx, name = "BRL/USD",
            line = list(color = clr_main, width = 2),
            hovertemplate = "%{x|%Y-%m}<br>BRL/USD: %{y:.3f}<extra></extra>") %>%
  layout(
    title = list(text = "<b>Simulated exchange rate (BRL/USD)</b><br><sub>Higher domestic rate → BRL appreciation (correct UIP sign)</sub>",
                 font = list(size = 12, color = clr_main)),
    xaxis = list(title = ""), yaxis = list(title = "BRL/USD"),
    hovermode = "x unified", plot_bgcolor="white", paper_bgcolor="white",
    legend = plt_layout$legend, font = plt_layout$font,
    margin = plt_layout$margin, hoverlabel = plt_layout$hoverlabel
  )

p_x <- plot_ly(df_uip, x = ~date) %>%
  add_lines(y = ~risk, name = "EMBI+",
            line = list(color = clr_red, width = 2),
            hovertemplate = "%{x|%Y-%m}<br>EMBI+: %{y:.0f} bps<extra></extra>") %>%
  layout(shapes = list(hline(200, clr_grey, "dot"))) %>%
  layout(
    title = list(text = "<b>Country risk premium (EMBI+)</b><br><sub>AR(1) with gradual fiscal improvement</sub>",
                 font = list(size = 12, color = clr_main)),
    xaxis = list(title = ""), yaxis = list(title = "Basis Points (bps)"),
    hovermode = "x unified", plot_bgcolor="white", paper_bgcolor="white",
    legend = plt_layout$legend, font = plt_layout$font,
    margin = plt_layout$margin, hoverlabel = plt_layout$hoverlabel
  )

subplot(p_e, p_x, nrows = 2, shareX = TRUE, titleY = TRUE) %>%
  layout(hovermode = "x unified") %>%
  config(plt_cfg)
```

---

# Block IV — Monetary Policy Rule

## Smoothed Taylor Rule

$$\boxed{i_t = (1-\lambda)\,i_{t-1} + \lambda \left[\alpha_1(\pi_t - \pi^*) + \alpha_2 \tilde{y}_t + \alpha_3\right]}$$

::::: {.warning-box}
**Normative vs. positive.** The original MPP used the rule normatively — to find the SELIC path consistent with target convergence under scenario X. Using it as a positive predictor of COPOM decisions is a different exercise. Label rule-implied paths as *"policy-consistent path under the estimated reaction function"*, not as SELIC forecasts.
:::::

```{r taylor-rules}
set.seed(55)
N_rules <- 80
alpha_pi_r <- 1.5; alpha_y_r <- 0.5; lambda_sm <- 0.7

i_taylor <- alpha_pi_r*(pi[1:N_rules]-pi_star) + alpha_y_r*df_gdp$gap[1:N_rules] + 0.10
i_smooth <- numeric(N_rules); i_smooth[1] <- i_taylor[1]
for (t in 2:N_rules) i_smooth[t] <- (1-lambda_sm)*i_smooth[t-1] + lambda_sm*i_taylor[t]
i_const <- rep(mean(i_taylor), N_rules)

df_rules <- data.frame(date = dates[1:N_rules],
                       Pure.Taylor = pmax(i_taylor, 0.03),
                       Smoothed    = pmax(i_smooth, 0.03),
                       Constant    = i_const)

plot_ly(df_rules, x = ~date) %>%
  add_lines(y = ~Pure.Taylor*100, name = "Pure Taylor",
            line = list(color = clr_main, width = 2),
            hovertemplate = "%{x|%Y-%m}<br>Pure Taylor: %{y:.1f}%<extra></extra>") %>%
  add_lines(y = ~Smoothed*100, name = "Smoothed Taylor (λ=0.7)",
            line = list(color = clr_red, width = 2),
            hovertemplate = "%{x|%Y-%m}<br>Smoothed: %{y:.1f}%<extra></extra>") %>%
  add_lines(y = ~Constant*100, name = "Constant Rate",
            line = list(color = clr_grey, width = 2, dash = "dash"),
            hovertemplate = "%{x|%Y-%m}<br>Constant: %{y:.1f}%<extra></extra>") %>%
  layout(
    title = list(
      text = "<b>Policy rate path under alternative rules</b><br><sub>Illustrative simulation — α_π = 1.5, α_y = 0.5</sub>",
      font = list(size = 13, color = clr_main)
    ),
    xaxis = list(title = ""),
    yaxis = list(title = "Policy Rate (% p.a.)", ticksuffix = "%"),
    hovermode = "x unified",
    plot_bgcolor = "white", paper_bgcolor = "white",
    legend = plt_layout$legend, font = plt_layout$font,
    margin = plt_layout$margin, hoverlabel = plt_layout$hoverlabel
  ) %>%
  config(plt_cfg)
```

---

# Full System Simulation

## Inflation Fan Chart

```{r fan-chart, fig.height=5}
set.seed(1234)
T_proj <- 12; N_sims <- 2000
a1_pc <- 0.45; a2_pc <- 0.20; a_fwd_pc <- 0.35   # sum = 1.00 ✓
a3_pc <- 0.25; a4_pc <- 0.12
b1_is <- 0.70; b2_is <- -0.15; b3_is <- -0.25
lambda_r <- 0.65; phi_pi <- 1.5; phi_y <- 0.5
pi0 <- 0.055; pi_1 <- 0.060; y0 <- -0.01; i0 <- 0.115; e0 <- log(5.0)

simulate_mpp <- function(seed_s) {
  set.seed(seed_s)
  pi_v <- c(pi_1, pi0, numeric(T_proj))
  y_v  <- c(0, y0, numeric(T_proj))
  i_v  <- c(i0, i0, numeric(T_proj))
  for (t in 3:(T_proj+2)) {
    r_real   <- i_v[t-1] - pi_v[t-1]
    y_v[t]   <- b1_is*y_v[t-1] + b2_is*y_v[t-2] + b3_is*r_real + rnorm(1,0,0.007)
    pi_fwd_t <- pi_v[t-1] + rnorm(1,0,0.004)
    delta_e_t <- -(i_v[t-1]-0.055-0.03)/4 + rnorm(1,0,0.015)
    pi_v[t]  <- a1_pc*pi_v[t-1] + a2_pc*pi_v[t-2] +
                a_fwd_pc*pi_fwd_t + a3_pc*y_v[t-1] +
                a4_pc*delta_e_t + rnorm(1,0,0.006)
    pi_v[t]  <- max(pi_v[t], 0)
    i_v[t]   <- (1-lambda_r)*i_v[t-1] +
                lambda_r*(phi_pi*(pi_v[t]-pi_star)+phi_y*y_v[t]+0.10)
  }
  pi_v[3:(T_proj+2)]
}

mat_pi <- sapply(1:N_sims, simulate_mpp)
qtis   <- as.data.frame(t(apply(mat_pi, 1, quantile,
                                probs = c(.10,.20,.30,.40,.50,.60,.70,.80,.90))))
colnames(qtis) <- paste0("q", c(10,20,30,40,50,60,70,80,90))
qtis$quarter   <- seq(as.Date("2024-04-01"), by="quarter", length.out=T_proj)

# ── Full plotly fan chart with interactive bands ────────────────────────────────
plot_ly(qtis, x = ~quarter) %>%

  # 80% band (p10–p90)
  add_ribbons(ymin = ~q10*100, ymax = ~q90*100,
              fillcolor = paste0(clr_main, "1A"),
              line = list(color = "transparent"),
              name = "80% band", legendgroup = "bands",
              hoverinfo = "skip") %>%

  # 60% band (p20–p80)
  add_ribbons(ymin = ~q20*100, ymax = ~q80*100,
              fillcolor = paste0(clr_main, "2A"),
              line = list(color = "transparent"),
              name = "60% band", legendgroup = "bands",
              hoverinfo = "skip") %>%

  # 40% band (p30–p70)
  add_ribbons(ymin = ~q30*100, ymax = ~q70*100,
              fillcolor = paste0(clr_main, "3A"),
              line = list(color = "transparent"),
              name = "40% band", legendgroup = "bands",
              hoverinfo = "skip") %>%

  # 20% band (p40–p60)
  add_ribbons(ymin = ~q40*100, ymax = ~q60*100,
              fillcolor = paste0(clr_main, "45"),
              line = list(color = "transparent"),
              name = "20% band", legendgroup = "bands",
              hoverinfo = "skip") %>%

  # Median line with rich tooltip
  add_lines(y = ~q50*100,
            name = "Median (p50)",
            line = list(color = clr_main, width = 3),
            hovertemplate = paste0(
              "<b>%{x|%Y-Q%q}</b><br>",
              "Median: %{y:.2f}%<br>",
              "<i>Hover bands for range</i><extra></extra>"
            )) %>%

  # p10 / p90 boundary lines (subtle)
  add_lines(y = ~q10*100, name = "p10",
            line = list(color = clr_main, width = 0.7, dash = "dot"),
            showlegend = FALSE,
            hovertemplate = "p10: %{y:.2f}%<extra></extra>") %>%
  add_lines(y = ~q90*100, name = "p90",
            line = list(color = clr_main, width = 0.7, dash = "dot"),
            showlegend = FALSE,
            hovertemplate = "p90: %{y:.2f}%<extra></extra>") %>%

  # Inflation target
  add_lines(y = rep(pi_star*100, T_proj),
            x = ~quarter,
            name = paste0("Target (", pi_star*100, "%)"),
            line = list(color = clr_red, width = 1.8, dash = "dash"),
            hovertemplate = "Target: %{y:.1f}%<extra></extra>") %>%

  layout(
    title = list(
      text = paste0("<b>Fan Chart — CPI Inflation Projection</b><br>",
                    "<sub>Monte Carlo (N=", N_sims, " paths) — 20/40/60/80% conditional forecast bands</sub>"),
      font = list(size = 14, color = clr_main)
    ),
    xaxis = list(title = "", tickformat = "%Y"),
    yaxis = list(title = "CPI (% p.a.)", ticksuffix = "%"),
    hovermode = "x unified",
    plot_bgcolor = "white", paper_bgcolor = "white",
    legend = list(orientation = "h", y = -0.18, font = list(size = 11)),
    font = plt_layout$font,
    margin = list(l=60, r=30, t=80, b=60),
    hoverlabel = plt_layout$hoverlabel
  ) %>%
  config(plt_cfg)
```

---

# Impulse Response: +200 bps Policy Shock

```{r irf-shock}
simulate_shock <- function(delta_selic = 0.02, seed_s = 42) {
  set.seed(seed_s)
  T_sim <- 16
  pi_v  <- c(0.060, 0.055, numeric(T_sim))
  y_v   <- c(0.005, -0.01, numeric(T_sim))
  i_v   <- c(0.115, 0.115 + delta_selic, numeric(T_sim))
  for (t in 3:(T_sim+2)) {
    r_real  <- i_v[t-1] - pi_v[t-1]
    y_v[t]  <- b1_is*y_v[t-1] + b2_is*y_v[t-2] + b3_is*r_real + rnorm(1,0,0.004)
    pi_fwd  <- pi_v[t-1] + rnorm(1,0,0.003)
    pi_v[t] <- a1_pc*pi_v[t-1] + a2_pc*pi_v[t-2] +
               a_fwd_pc*pi_fwd + a3_pc*y_v[t-1] + rnorm(1,0,0.004)
    pi_v[t] <- max(pi_v[t], 0)
    i_v[t]  <- i_v[2]
  }
  data.frame(t = 1:T_sim,
             inflation = pi_v[3:(T_sim+2)] * 100,
             gap       = y_v[3:(T_sim+2)] * 100,
             selic     = i_v[3:(T_sim+2)] * 100,
             scenario  = ifelse(delta_selic > 0, "+200bps shock", "Baseline"))
}

df_irf <- bind_rows(simulate_shock(0.00), simulate_shock(0.02))

make_irf_line <- function(y_var, label_var, y_title, add_target = FALSE) {
  df_base  <- df_irf %>% filter(scenario == "Baseline")
  df_shock <- df_irf %>% filter(scenario == "+200bps shock")
  p <- plot_ly() %>%
    add_lines(data = df_base,
              x = ~t, y = ~get(y_var),
              name = "Baseline",
              line = list(color = clr_grey, width = 2, dash = "dot"),
              hovertemplate = paste0("Q%{x}<br>Baseline ", label_var, ": %{y:.2f}%<extra></extra>")) %>%
    add_lines(data = df_shock,
              x = ~t, y = ~get(y_var),
              name = "+200bps",
              line = list(color = clr_red, width = 2.5),
              hovertemplate = paste0("Q%{x}<br>+200bps ", label_var, ": %{y:.2f}%<extra></extra>"))
  if (add_target)
    p <- p %>% layout(shapes = list(hline(pi_star*100, clr_main, "dot", 1)))
  p %>% layout(
    title  = list(text = paste0("<b>", y_title, "</b>"),
                  font = list(size = 12, color = clr_main)),
    xaxis  = list(title = "Quarters", dtick = 2),
    yaxis  = list(title = paste0(y_title, " (%)"), ticksuffix = "%"),
    hovermode = "x unified",
    plot_bgcolor = "white", paper_bgcolor = "white",
    legend = plt_layout$legend, font = plt_layout$font, hoverlabel = plt_layout$hoverlabel
  )
}

p_inf  <- make_irf_line("inflation", "Inflation", "Inflation",    add_target = TRUE)
p_gap2 <- make_irf_line("gap",       "Gap",       "Output gap",   add_target = FALSE)
p_sel  <- make_irf_line("selic",     "Selic",     "Policy rate",  add_target = FALSE)

subplot(
  subplot(p_inf, p_gap2, shareX = TRUE, titleY = TRUE),
  p_sel,
  nrows = 2, shareX = TRUE, titleY = TRUE
) %>%
  layout(
    title = list(
      text = "<b>Impulse Response — Contractionary Policy Shock (+200 bps)</b><br><sub>Disinflation via demand channel with ~2–3 quarter lag</sub>",
      font = list(size = 13, color = clr_main)
    ),
    hovermode = "x unified",
    plot_bgcolor = "white", paper_bgcolor = "white"
  ) %>%
  config(plt_cfg)
```

---

# Limitations

```{r limitations-table, echo=FALSE}
df_lim <- data.frame(
  Issue = c(
    "Long-run restriction not imposed (Phillips)",
    "Overlapping 12m inflation variable",
    "UIP not fully identified",
    "Neutral rate is a statistical trend",
    "OLS with no simultaneity correction",
    "Monthly frequency (vs. original quarterly)"
  ),
  Description = c(
    "OLS does not enforce α₁+α₂+α_fwd=1; holds in simulation but not estimation",
    "12m CPI creates mechanical autocorrelation; monthly or non-overlapping measures preferred",
    "Estimated as VAR/OLS auxiliary; foreign rate proxied rather than measured",
    "r* estimated from rolling ex-ante real rates — reflects policy reactions, not structural equilibrium",
    "IS, Phillips, and UIP are simultaneously determined; OLS does not identify structural channels",
    "IBC-Br is a reasonable proxy but introduces differences vs. quarterly GDP"
  ),
  Fix = c(
    "Impose as linear constraint in nlm() or systemfit",
    "Use monthly or trimestral non-overlapping inflation",
    "Use a mapped foreign rate series; impose UIP sign restriction",
    "Report r* scenarios: {4%, 5%, 6%}",
    "Instrument with lagged values or use SVAR identification",
    "Acknowledge frequency adaptation; verify lag structure in months"
  )
)
knitr::kable(df_lim, col.names = c("Issue","Description","Recommended Fix"),
             caption = "**Known limitations and recommended corrections**") %>%
  kableExtra::kable_styling(bootstrap_options=c("striped","hover","condensed")) %>%
  kableExtra::column_spec(1, bold=TRUE, color=clr_main) %>%
  kableExtra::column_spec(3, italic=TRUE, color="#457b9d")
```

---

# References

- Bogdanski, J., Tombini, A.A. & Werlang, S.R.C. (2000). *Implementing Inflation Targeting in Brazil*. Working Paper 1, Banco Central do Brasil.
- Goldfajn, I. & Werlang, S.R.C. (1999). *The Pass-through from Depreciation to Inflation*. Mimeo, PUC-Rio / BCB.
- Taylor, J.B. (1993). Discretion versus Policy Rules in Practice. *Carnegie-Rochester Conference Series on Public Policy*, 39, 195–214.
- Hodrick, R.J. & Prescott, E.C. (1997). Postwar U.S. Business Cycles. *Journal of Money, Credit and Banking*, 29(1), 1–16.
- Castro, M.R. et al. (2011). *SAMBA: Stochastic Analytical Model with a Bayesian Approach*. Working Paper 239, BCB.

---

*All data are **simulated** for pedagogical purposes. This document is an adaptation — not a replication — of the BCB Small-Scale Model.*
